---
author: "James McDermott"
institute: "NUI Galway"
title: "CT5132/CT5148 Week 12 Exercises"
output:
  beamer_presentation:
    colortheme: "rose"
    fonttheme: "structurebold"
    theme: "Madrid"
    slide_level: 2
    fig_caption: false
    includes:
      in_header: beamer-header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Exercises

Again this week, our exercises are extracted from the lecture slides/videos, and solutions are given below.

We won't look at the exercises for `ggplot`, as the Data Visualisation module is coming up soon where you'll see that in much more detail.

## Exercises (`dplyr` joins)

1. Read the three data files `rentals.csv`, `movies.csv`, `customers.csv`, all in the `data/` directory, as tibbles.
2. Optional: get R to read the Date column correctly. Hint: https://readr.tidyverse.org/reference/parse_datetime.html
3. Using a `dplyr` join command, create a table showing the customer name and address for every rental.
4. Piping the result into another join command, recreate the full original table as shown under "Before Normalisation" above.
5. Notice the columns `Name.x` and `Name.y` which appear because there is a `Name` column in each of the Movies and Customers tables. Rename them.
6. Calculate the number of movies Frida watched of the Sci-fi genre.

## Solutions (`dplyr` joins)

```{r}
library(tidyverse)
```


## Exercises 1 and 2:
```{r}
rentals <- read_csv("data/rentals.csv", 
                    col_types=cols(Date=col_date(
                      format="%d-%b-%Y")))
movies <- read_csv("data/movies.csv")
customers <- read_csv("data/customers.csv")
```

## Customer name and address for each rental
```{r}
inner_join(rentals, customers, by="CustomerID")
```

## Recreate original table
```{r}
inner_join(rentals, customers, by="CustomerID") %>% 
  inner_join(movies, by="MovieID")
```

## Rename columns

```{r}
t = inner_join(rentals, customers, by="CustomerID") %>% 
  inner_join(movies, by="MovieID") %>% 
  rename(CustomerName=Name.x, MovieTitle=Name.y)
t
```

## Filter and count

```{r}
t %>% filter(CustomerName=="Frida", Genre=="Sci-fi") %>% 
  count()
```

## Filter and count

The following is a solution to the problem, but it requires the programmer to do all the work in their head. That's not scalable or flexible and it's error-prone, so don't do this.
```{r}
# Frida is CustomerID 2
# Movies 101 and 104 are Sci-fi
rentals %>% filter(CustomerID == 2, 
                   MovieID %in% c(101, 104)) %>%
  count()
```




## Exercises

1. In the `mpg` dataset (part of the tidyverse), calculate the mean and standard deviation of the highway fuel efficiency.
2. Using `group_by`, calculate the mean and standard deviation of the highway fuel efficiency per manufacturer.
3. Calculate the correlation between highway fuel efficiency and engine size.
4. What was the average highway fuel efficiency in 1999 and in 2008?
5. Carry out a two-sample independent t-test between highway fuel efficiency in 1999 and 2008 and interpret the result.
6. Carry out a regression on highway fuel efficency by displacement.

## Solution 1
library(tidyverse)

```{r}
mean(mpg$hwy)
sd(mpg$hwy)
```

## Solution 2
```{r}
mpg %>% group_by(manufacturer) %>% 
  summarise(mean=mean(hwy), sd=sd(hwy))
```

## Solution 3

```{r}
cor(mpg$hwy, mpg$displ)
```

## Solution 4
```{r}
mpg %>% group_by(year) %>% 
  summarise(mean=mean(hwy), sd=sd(hwy))
```

## Solution 5
```{r}
mpg1999 <- mpg %>% filter(year == 1999)
mpg2008 <- mpg %>% filter(year == 2008)
t.test(mpg1999$hwy, mpg2008$hwy)
```

## Solution 6

```{r}
res = lm(hwy ~ displ, data=mpg)
summary(res)
```






