---
author: "James McDermott"
institute: "NUI Galway"
title: "CT5132/CT5148 Week 11 Exercises"
output:
  beamer_presentation:
    colortheme: "rose"
    fonttheme: "structurebold"
    theme: "Madrid"
    slide_level: 2
    fig_caption: false
    includes:
      in_header: beamer-header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Week 11 Exercises

The exercises and solutions are extracted from the lecture slides.

```{r}
library(tidyverse)
```


## Exercises (R Basics)

1. Write the Factorial function in R, eg `fact(5)` gives `120`.
2. Given `x <- "John"`, calculate the length in characters of `x`. Use `nchar()`.
3. Given `xs <- c("John", "Paul", "George", "Ringo")`,
   calculate the length of each name, using vectorisation (not a for-loop).
4. Calculate whether each name is shorter than 5 characters.
5. Index `xs` to keep just the names shorter than 5 characters.
6. Write a function which unit-norms a vector, ie normalises it so that the vector length equals 1.
   Eg `unit_norm(c(10, 10, 10, 10))` gives `0.5 0.5 0.5 0.5`.
7. Write a function which standardises a vector, ie gets the $z$-score, ie maps it to have mean 0 and standard deviation 1.
Eg `z_score(c(10, 6, 12, 12))` gives `0.0000000 -1.4142136  0.7071068  0.7071068`.




## Solutions (R Basics)

```{r}
fact <- function(n) { # Exercise 1
  if (n <= 1) {
    1 # remember, no return statement!
  } else {
    n * fact(n-1)
  }
}
fact(5)
```

***

```{r}
x <- "John"
nchar(x) # Exercise 2
xs <- c("John", "Paul", "George", "Ringo")
nchar(xs) # Exercise 3
nchar(xs) < 5 # Exercise 4
xs[nchar(xs) < 5] # Exercise 5
```

*** 

```{r}
unit_norm <- function(x) { # Exercise 6
  x / sqrt(sum(x**2))
}
unit_norm(c(10, 10, 10, 10))
```
  
```{r}
z_score <- function(x) { # Exercise 7
  (x - mean(x)) / sd(x)
}
z_score(c(10, 6, 12, 12))
```


## Exercises (Tidy Data)

1. Recall our experiment on running time for sorting an array of different sizes. The original data (before we added extra columns) is available in `data/sort_times_original.csv`. Read it in to a `tibble`. (You might need to set the working directory first.)
2. Use `glimpse` to take a look. What types do the columns have?
3. In what way is this *not* tidy data? Use `gather` to fix it. Hint: the result should have shape 50 x 3 with columns `n`, `run_number`, `run_time`.
4. It would be nicer if `run_number` was just an integer, eg `0`, instead of `run0`. Use `separate` to split it into two parts. Hint: use `into=c("dummy", "run_number")`. 
5. Look again at the result. We don't need that "dummy" column. Use `NA` to omit it. Hint: see `?separate` for help on `into`.
6. Look again -- `run_number` is still not an integer! Fix this. Hint: `separate` can guess the correct type to convert to, but see `?separate` again to see how to ask it to.
7. Write it to a file `data/sort_times_tidy.csv` using `write_csv()`.

## Solutions (Tidy Data)

## Exercise 1

```{r}
d <- read_csv("data/sort_times_original.csv") 
```

## Exercise 2

```{r}
glimpse(d) # All columns of type `dbl`, which is ok
```

## Exercise 3

```{r}
d <- gather(d, key="run_number", value="run_time", 
            run0, run1, run2, run3, run4) 
```

## Exercise 4

```{r}
separate(d, run_number, 
         into=c("dummy", "run_number"), sep=3) 
```

## Exercise 5

```{r}
separate(d, run_number, 
         into=c(NA, "run_number"), sep=3) 
```

## Exercise 6

```{r}
d <- separate(d, run_number, 
              into=c(NA, "run_number"), sep=3, 
              convert=TRUE)
d
```

## Exercise 7
```{r}
write_csv(d, "data/sort_times_tidy.csv")
```




## Exercises (`dplyr`)

* Exercise 1: Our sort times data is available in tidy format as `sort_times_tidy.csv`. Use `group_by` and `summarise` to get the mean and the standard deviation for each `n`, and then for each `run_number`. 

* A dataset of characters in *Star Wars* is available as `dplyr::starwars`. Exercise 2: Find all the human females. Exercise 3: Find the characters who are human *or* Wookiee. Exercise 4: Find the shortest character. Hint: recall we might need `na.rm`. Exercise 5: Add a new column called `BMI` giving the body mass index, where the formula is BMI = $m / h^2$ for mass $m$ in kg and height $h$ in metres. https://en.wikipedia.org/wiki/Body_mass_index. Exercise 6: Which character has the highest BMI?

## Solutions (`dplyr`)

## Exercise 1

```{r}
d <- read_csv("data/sort_times_tidy.csv")
d %>% group_by(n) %>% 
  summarise(mean_run_time=mean(run_time), 
            sd_run_time=sd(run_time))
```

## Exercise 1

Notice that the mean and stddev for $n=7$ million are anomalously high. One way this could occur is if our computer had a spike in CPU usage during the experiment, e.g. due to a browser loading a video.

## Exercise 1
```{r}
d %>% group_by(run_number) %>% 
  summarise(mean_run_time=mean(run_time), 
            sd_run_time=sd(run_time))
```

No major anomalies this time.


## Exercise 2

```{r}
sw <- dplyr::starwars
# human females
sw %>% filter(species == "Human", gender == "female") 
```

## Exercise 3

```{r}
# human or Wookiee
sw %>% filter(species == "Human" | species == "Wookiee") 
# sw %>% filter(species %in% c("Human", "Wookiee")) # alternative
```

## Exercise 4

```{r}
sw %>% filter(height == max(height, na.rm=TRUE))
```


## Exercise 5
```{r}
# NB convert height from cm to metres before squaring
BMI <- function(h, m) {m / (h / 100)^2}
sw <- sw %>% mutate(bmi=BMI(height, mass))
```

## Exercise 6
```{r}
sw %>% filter(bmi == max(bmi, na.rm=TRUE))
```
